---
title: "Trends in Ecology Jobs by Analyzing Ecolog Posts With R"
author: "Sheng Pei Wang"
date: "Nov 13, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
# Why and what am I doing here?
## Back story
I have subscribed for the Ecolog listserv for a long time, hoping for the perfect position just for me. But given the growing competition for the stagnating job market in academia, I decided to plunge into the "real world" to pursue my interests in understanding the world through thinking, experimentation, and analysis. Before I leave my academic evolutionary ecologist dream behind, I want to figure out whether this one source of data supports the broader trend in the academic job market. Specifically, I want to address these three questions:   

## Questions
1. Is there an oversupply? In other words, are lower level positions more abundant than higher level positions. If so, I should observe number of posts for PhD > postdoc > faculty    
2. Is there growing supply of PhD over time? This should show up as an increase in numbers of PhD or postdoc positions.     
3. On the other hand, is the job market stagnating? This would be represented by faculty level positions being constant over time?

## Methods
All ecolog's posts are available online, dating back to 2000. This makes this a great data source, but the amount of data is not trivial to process. I decided to focus only on the titles of the posts, because it is usually easy to tell what audience the posts target.   
To get the data, I need to **scrap the website** for titles of every posts and to decide the target level (PhD/postdoc/faculty) of each post using **text mining**.

# Part 1. Getting the data

###Set up the R environment:
```{r, message = F}
library(rvest)    # Parsing of HTML/XML files 
library(tidyverse)  # General-purpose data wrangling
library(stringr)    # String manipulation

library(zoo)        # for the yearmon class
library(lme4)
```

## Creating the list of webpages to scrap

The actual R code file named "Fun Ecolog Data Scrapping Project.R" has the codes with which I explored how to get the data. But the simple story is this:

I notice that the posts were stored on different pages grouped by weeks, and the master list is at https://listserv.umd.edu/archives/ecolog-l.html. So I can scrap the master list to get the weekly lists' url's, and then I can scrape the weekly lists to get the titles of every post:


```{r message=FALSE, warning=FALSE}
page_archive <- read_html("https://listserv.umd.edu/archives/ecolog-l.html")
# Some times this command end up with an error message of "Timeout", because the archive's server is picky about the connections. My home network tend to do this. The only advice I have is to be patient and maybe try a different network if possible.

# I noticed that by interacting with html source code of the full list, all the weekly links are stored under the "li" node
pages_info  <- page_archive %>%  html_nodes("li")  # Scrap the weekly link info from main page

list_of_weeks <- page_archive %>%  html_nodes("li") %>% html_text()
list_of_weeks <- list_of_weeks[-c(1:4)]

urls <- lapply(pages_info, function(x) str_sub(x,15, 52)) # get only the url part, which is between 15-52 characters.
urls <- urls[-c(1:4)] # remove the first 4 items that are not related to our purpose.

# Making the list of urls
list_of_pages <- str_c('https://listserv.umd.edu', urls)

head(list_of_pages)
```

## A function to get the relevant information 
### (titles posted in the last 20 years)
To  actually get the data from the urls, I wrote a small functions to get the necessary information from each page. Then apply the function to the list of urls.
```{r}
#I need to create a custom function that gets all postings from each page
ecolog_p <- function(url){
  webpage <- read_html(url)
  Sys.sleep(0.2)
  webpage %>% html_nodes("tr:nth-child(3) a") %>%
              html_text
}

# ecolog_postings<- lapply(list_of_pages, ecolog_p)
```

I commented out the last line of code so it's not run here, because it takes about 2 hours to get all the data with a good internet connections. If anyone were to run this line of code, I recommend breaking it down into chunks.

# Part 2. Text mining of the tiles
The idea behind this section is simple: I want to determine if a post is advertising a PhD, postdoc, or faculty position. 


## A function to characterize the type of positions from only titles
Characterizing the target of a post can be complicated, but I don't need to be precise about every post here given the amount of posts (more than 10,000 across over 18 years). So I took a simple approach of counting the occurrence of key words such as PhD to determine the abundance of each level of posts. By reading through a few months' worth of posts myself, I confirmed that the key words are good indicator of the content of the posts, and that they are rarely repeated in a single title.
```{r}
#I wrote another function to count the appearance of different key words.
position_breakdown <- function(x){
  x = str_to_lower(x)
  u = sum(str_count(x, "reu|intern|volunteer"))
  m = sum(str_count(x, "ms |m.s |msc |m.sc "))  
  phd = sum(str_count(x, "phd"))
  po = sum(str_count(x, "postdoc|post-doc|post doc"))  
  f = sum(str_count(x, "director|faculty|professor|instructor"))
  o = sum(str_count(x, "technician|assistant|manager"))
  l = length(x)
  y = data.frame(undergrad = u, ms = m, phd = phd, postdoc = po, 
                faculty = f, otherjob = o, total = l)
  y}

```
## Get a small sample dataset:
I will be using a sample page from the first week of 2018:
```{r}
url <- "https://listserv.umd.edu/cgi-bin/wa?A1=ind1804a&L=ecolog-l"
webpage <- read_html(url)
page_4i18 <- html_text(html_nodes(webpage,"tr:nth-child(3) a")) #First download the webpage
# A quick note that I used the web tool "Selector Gadget"" to generate the CSS selectors. i.e. the ""tr:nth-child(3) a"" part
```

## Look at data using the sample dataset:
```{r}
# An example of using the position_breakdown function.
position_breakdown(page_4i18)
```

As for this particular week, you can see their are increasing numbers of jobs with increasing training, from undergrad to master to PhD to postdoc positions. However, this trend stops at the postdoc level, and there are only 4 faculty positions. **The number of postdoc positions is 3.5 times that of faculty positions in this one week, supporting my first hypothesis that there is a oversupply of PhDs** I will analyze the full dataset in the next section to test whether this trend holds.


# Part 3. Analyzing the data
## Data cleanup
I will be brief here. I scraped all the webpages available up until now (11/25/18) and saved the list as "data_ecolog_post_Nov_2018.R". And I will measure the number of each type of postings as explained above. I also added the time information. Then I summarized each month's data as a single entry, because the fourth week of a month is often less than 7 days, which will skew the analyses.
```{r}
load("data_ecolog_post_Nov_2018.R")
data_positions <- lapply(data_ecolog_post_updated, position_breakdown)

list_of_month <- 
  sapply(list_of_weeks, function(x) strsplit(x, ",")[[1]]) %>% 
  matrix(ncol=2, byrow=T)

list_of_months <- list_of_month[,1] %>% 
  as.yearmon()

data_jobs <- 
  do.call(rbind.data.frame, data_positions) %>% 
  mutate(month = list_of_months)
head(data_jobs)

data_jobs_month <-
  data_jobs %>% 
  group_by(month) %>% 
  summarise_all(sum) 
head(data_jobs_month)

# reshaping the dataset into the long format for visualization and analysis
data_jobs_month_long <- 
  data_jobs_month %>% 
  gather(level, number_of_posts, -month) %>% 
  mutate(level = fct_recode(as.factor(level), 
                            Faculty = "faculty",
                            Postdoc = "postdoc",
                            PhD = "phd",
                            MS = "ms",
                            Undergrad = "undergrad",
                            other_job = "otherjob"),
         level = fct_relevel(level,
                             "Faculty",
                             "Postdoc",
                             "PhD",
                             "MS",
                             "Undergrad",
                             "other_job"))
```

## Data visualization
Let's first look at the the data for all the levels
```{r}
ggplot(data_jobs_month_long %>% filter(level != "total"), 
       aes(month, number_of_posts, color = level)) +
  geom_point() +
  geom_smooth(method = 'loess') +
  scale_colour_hue(c = 200) +
  theme_classic()
```
There is a pretty clear trend of increase over time across all levels besides undergrad level. **Faculty level jobs increased less compare to PhD or postdoc level jobs.** There are more postings about other types of jobs, and the growth in that section is rapid. 

Is the overall trend of increase real? Could it be that ecolog is becoming more popular over time?
```{r}
ggplot(data_jobs_month_long %>% filter(level == "total"), 
       aes(month, number_of_posts)) +
  geom_point() +
  geom_smooth(method = 'loess') +
  theme_classic()
```
More people are posting on ecolog now than 15 years ago, so ecolog is getting more popular. There is a slight dip around 2004, and I would like to explore this more later. But I don't know if this is the driving force behind the increase in job postings. 

I will explore the proportional changes in job postings to understand whether the increase in job postings is driven by the growing popularity of ecolog.
First need to convert the count data into proportions:
```{r}
data_jobs_proportions <-
  data_jobs_month %>% 
  mutate(other = undergrad + phd + postdoc + faculty + otherjob,
         other = total - other,
         other = other/total,
         undergrad = undergrad/total,
         ms = ms/total,
         phd = phd/total,
         postdoc = postdoc/total,
         faculty = faculty/total,
         otherjob = otherjob/total)
  
data_jobs_proportions_long <- 
  data_jobs_proportions %>% 
  gather(level, proportion_of_posts, -month) %>% 
  mutate(level = fct_recode(as.factor(level), 
                            Faculty = "faculty",
                            Postdoc = "postdoc",
                            PhD = "phd",
                            MS = "ms",
                            Undergrad = "undergrad",
                            other_job = "otherjob"),
         level = fct_relevel(level,
                             "Faculty",
                             "Postdoc",
                             "PhD",
                             "MS",
                             "Undergrad",
                             "other_job"))
```
The code here can be streamlined, and I would welcome any comments.

Let's first plot all the proportional data:
```{r}
ggplot(data_jobs_proportions_long %>% filter(level != "total"), 
       aes(month, proportion_of_posts, color = level)) +
  geom_point() +
  geom_smooth(method = 'loess') +
  scale_colour_hue(c = 200) +
  theme_classic()
```
This is interesting, that the proportion of job-posts has increased over time. Even though most of the posts are not clearly job related, the number of non-job-related postings is in stark decline in terms of proportions.
**So ecolog has become more popular over time, but it's popularity as a job board is growing even faster.**

This figure does not show how the job posting proportions have changed, so let's look at just job-related posts.
```{r}
ggplot(data_jobs_proportions_long %>% 
         filter(level %in% c("Faculty",
                             "Postdoc",
                             "PhD",
                             "MS",
                             "Undergrad",
                             "other_job")), 
       aes(month, proportion_of_posts, color = level)) +
  geom_point() +
  geom_smooth(method = 'loess') +
  scale_colour_hue(c = 200) +
  theme_classic()
```
Again, the increasing proportion of nearly all types of job posts is interesting. 

**The results partially support my first hypothesis that there is an oversupply of labor into the academic job market in ecology. Over the entire 18 years of existence of ecolog, the number of posts advertising for postdoc positions has been consistently higher than for faculty positions.** However, advertising for PhD and master student positions has been lower than for faculty positions. This is by no means a new finding, but it is interesting to see it represented in ecolog job posts. 

*The above conlusions have to be taken with a grain of salt though, given that the data is solely about ecolog. From my experience, advertising for student positions is more limited to lab webpage and twitter than for postdocs or faculty, the latter usually requires a broader search. So I don't think the pattern of "undersupply" of students shown in ecolog's posts represents the broader pattern.* 



## Formal analysis

I first altered the month data a little, so the mean and the std is close to each other. This will allow the models to converge and has no impact on the statistical implications.
```{r}
data_jobs_fit <-
  data_jobs_month_long %>% 
  filter(level != "total") %>% 
  mutate(month_n = as.numeric(month)-2000)
```

Fit a linear mixed model to understand the effects of time on each level. I specified that both the intercept and slopes can vary for each job post level.
```{r}
fit_time <- lmer(number_of_posts ~ month_n + (1 + month_n|level),
                 data = data_jobs_fit)

summary(fit_time)
```

The coefficient for the fixed effect is 1.66 with t value = 5.4, Suggesting an overall increase of number of posts over time. This is consistent with the figures.
```{r}
fit_time_null <- lmer(number_of_posts ~ (1 |level),
                 data = data_jobs_fit)
anova(fit_time_null, fit_time)
```
I used a model comparison with the null model to test the significance of the increase, and the statistical support for the increasing trend is very strong.

Let's then look at the different levels:
```{r}
ranef(fit_time) 
```
These are the random effects, basically the difference among the different levels. I'm particularly interested in the slope coefficients (month_n column). A few conclusions can be drawn:   
    + Increase in "other jobs"" is the fastest.
    + **The number of faculty positions closely followed the overall trend, but increase in postdoc positions is much faster than faculty positions.**
    + The increase in MS positions is the slowest. Notice that the actual slope would be -0.86 + 1.66, which is still positive.

**These results support my hypotheses 2 and 3:**
    + There is an increase in labor supply in terms of both PhD and postdocs, but the over-supply of postdoc is a lot stronger.
    + The job market defined by faculty job availability is stagnating in terms of closely matching the overall trend of job postings over time.

